---
title: "Homework 1"
output:
  html_document: default
  html_notebook: default
---
## Problem 1

### 1.1
Read `health.csv` and check basic information:
```{r}
health <- read.csv("health.csv")
head(health)
```
```{r}
str(health)
```

### 1.2
12610 observations came from `Female`, and 11805 observations came from `Male`. There are also 247 missing values for the `sex` column.
```{r}
summary(health$sex)
```

### 1.3
2187 missing values for the variable `height`.
```{r}
sum(is.na(health$height))
```

For those non-missing values, 29 of them are larger than 1.875.
```{r}
sum(health$height>1.875, na.rm=TRUE)
```

### 1.4
Levels for the variable `hands_eating`:
```{r}
levels(health$hands_eating)
```

Re-arrange the levels in the right order:
```{r}
health$hands_eating <- factor(health$hands_eating, levels=c("Always","Most of the time","Sometimes","Rarely","Never"))
levels(health$hands_eating)
```

## Problem 2
### 2.1
Read `unemployment.txt`:
```{r}
unemployment <- read.table("unemployment.txt", head=TRUE, sep="|")
```

### 2.2
For the month of Aug-11, Imperical (CA), Yuma (AZ) and Salinas (PR) are three counties with the highest unemployment rates.
```{r}
library(dplyr)
countySortedByRate <- unemployment %>% filter(period == "Aug-11") %>% arrange(desc(rate)) %>% select(county,state, rate)
head(countySortedByRate, 3)
```

Mckenzie, Slope and Williams are three counties with the lowest unemployment rates. All of them are in North Dakota.
```{r}
tail(countySortedByRate, 3)
```

### 2.3
For the month of Aug-11, Whitley is the county in Indiana that has unemployment rate closest to the overall unemployment rate across the whole country.
```{r}
overallRate <- with(unemployment, sum(unemployed)/sum(as.numeric(civilian))*100)
unemployment %>% filter(period == "Aug-11", state_code == "IN") %>% mutate(diff = abs(rate-overallRate)) %>% filter(diff == min(diff)) %>% select(county, rate, diff)
```

## Problem 3
### 3.1
Read `microarray.txt` into R, the dimension of the data frame is 6830 (genes) x 64 (cell lines).
```{r}
dat <- read.table("microarray.txt", head=TRUE)
dim(dat)
```

### 3.2
`bvar` takes as inpute a data frame and a vector of row numbers and returns which of these rows has the largest standard deviation.
```{r}
bvar <- function(data, rows) {
  rowSd <- apply(data[rows,], 1, sd)
  rows[which.max(rowSd)]
}
bvar(dat, c(40, 50, 60))
```

### 3.3
Plot heatmap of samples with label LEUKEMIA or MELANOMA.
```{r}
dat <- dat %>% select(starts_with("LEUKEMIA"), starts_with("MELANOMA"))
heatmap(t(as.matrix(dat)), labCol=FALSE, scale="none")
```

### 3.4
Replot the heatmap in red-green color system.
```{r}
library("gplots")
heatmap.2(t(as.matrix(dat)), col=greenred, breaks=-2.5:2.5, labCol=FALSE, margins=c(0,10), scale="none", trace="none", density.info="none")
```